---
title: "Famale Labor Participation in STEM Areas"
date: "2024-05-27"
output: rmarkdown::github_document
---

This repository provides an analysis of female labor participation in STEM areas for the New Dimensions program.

New Dimensions is a comprehensive business intelligence program for women from disadvantaged backgrounds in Ecuador. Our goal is to open pathways for Ecuadorian women in STEM through data education and employability skills. This analysis provides context and background for the need for the program.

The inaugural cohort included 55 women and received $25,000 in funding from the US Embassy in Ecuador in 2023. We are now seeking support for the upcoming second cohort scheduled to launch in 2025. Our goal is to secure funding for the next four cohorts (i.e. 2025 through 2028).

Elaborated by: Alonso Quijano [X](https://x.com/aquijanoruiz), [Github](https://github.com/aquijanoruiz)  
Source: Ecuador 2023 National Labor Survey [ENEMDU](https://www.ecuadorencifras.gob.ec/enemdu-anual/)

## Data preprocessing

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
  

```{r packages and data, results = 'hide', message=FALSE, warning=FALSE}
# -------------------------------------------------- #
# (1) Loading the data
# -------------------------------------------------- #

# Load libraries
library(haven)
library(labelled)
library(tidyverse)
library(survey)

# Load data
# Set temporary directory
td <- tempdir()

# Unzip the data file
unzip(zipfile = "ENEMDU_ANUAL_2023.zip", exdir = td)

# Read the SPSS file
enemdu2023 <- read_spss(file.path(td, "BDDenemdu_personas_2023_anual.sav"))
enemdu2023_labels <- var_label(enemdu2023)

```

```{r preprocessing}
# -------------------------------------------------- #
# (2) Preprocessing and Data Manipulation
# -------------------------------------------------- #

# Define economically active population (PEA)
enemdu2023 <- enemdu2023 %>% mutate(pea = ifelse(between(condact, 1, 8), 1, 0))

# Define adequate employment (people earning the minimum salary or higher)
enemdu2023 <- enemdu2023 %>% mutate(emp_adeq = ifelse(condact == 1, 1, 0))

# Define IESS contribution (1 if contributes to IESS, 0 otherwise)
enemdu2023 <- enemdu2023 %>% 
  mutate(iess = ifelse(between(p05a, 1, 4) | between(p05b, 1, 4), 1, 0))

# Define STEM field (1 if in STEM, 0 otherwise)
# We exclude people in health sciences to avoid overrepresentation of women in nursing and midwifery
enemdu2023$stem <- ifelse(substr(enemdu2023$p41, 1,2) %in% c(21, 25), 1, 0)

# Remove non responses from income variable
enemdu2023 <-  enemdu2023 %>% mutate(
  income = case_when(ingrl == -1 ~ 0, ingrl == 999999 ~ NA, TRUE ~ ingrl))
```

## Analysis w/o survey design

```{r analysis w/o survey weights}
# -------------------------------------------------- #
# (3) Employment Classification
# -------------------------------------------------- #

# Men
enemdu2023 %>% filter(p02 == 1, pea == 1) %>% 
  group_by(condact) %>% summarize(n = n()) %>% 
  mutate(prob = prop.table(n))

# Women
enemdu2023 %>% filter(p02 == 2, pea == 1) %>% 
  count(condact) %>% mutate(prob = prop.table(n))

# Indigenous Women
enemdu2023 %>% filter(p02 == 2, pea == 1, p15 == 1) %>% 
  count(condact) %>% mutate(prob = prop.table(n))

# -------------------------------------------------- #
# (4) Occupation Type
# -------------------------------------------------- #

# Men
enemdu2023 %>% filter(p02 == 1, empleo == 1) %>% 
  count(grupo1) %>% mutate(prob = prop.table(n))

# Women
enemdu2023 %>% filter(p02 == 2, empleo == 1) %>%
  count(grupo1) %>% mutate(prob = prop.table(n))

# Indigenous women
enemdu2023 %>% filter(p02 == 2, empleo == 1, p15 == 1) %>%
  count(grupo1) %>% mutate(prob = prop.table(n))

# -------------------------------------------------- #
# (5) Income
# -------------------------------------------------- #

# Average income by gender (STEM only)
enemdu2023 %>% filter(stem == 1) %>% group_by(as_factor(p02)) %>% 
  summarize(avg_income = mean(income, na.rm = TRUE))

# Median income by gender (STEM only)
enemdu2023 %>% filter(stem == 1) %>% group_by(as_factor(p02)) %>% 
  summarize(avg_income = median(income, na.rm = TRUE))

```

## Analysis w/ survey design

```{r analysis w/ survey weights}
# -------------------------------------------------- #
# (6) Survey Design
# -------------------------------------------------- #

# Create survey design object
enemdu2023_svydesign <- svydesign(data = enemdu2023,
                                  ids = ~ upm, # Primary Sampling Unit
                                  strata = ~ estrato, # Strata
                                  weights = ~ fexp, # Weights
                                  nest = TRUE) # Multistage Sampling

# -------------------------------------------------- #
# (7) Analysis with Survey Design
# -------------------------------------------------- #

# Employment classification of economically active women
svymean(x = ~ as_factor(condact),
        design = subset(enemdu2023_svydesign, pea == 1 & p02 == 2), 
        na.rm = TRUE)

# Percentage of women who contribute to IESS
svymean(x = ~ iess,
        design = subset(enemdu2023_svydesign, pea == 1 & p02 == 2), 
        na.rm = TRUE)

# Gender classification of individuals in STEM fields
svymean(x = ~ as_factor(p02),
        design = subset(enemdu2023_svydesign, stem == 1), 
        na.rm = TRUE)

# Educational attainment of adult women
svymean(x = ~ as_factor(nnivins),
        design = subset(enemdu2023_svydesign, pea == 1 & p02 == 2 & p03 >= 18), 
        na.rm = TRUE)

# Average income by gender
svyby(~ income, ~ as_factor(p02), 
      design = enemdu2023_svydesign, 
      svymean, na.rm = TRUE)

(419.7013 - 495.6362) / 495.6362

# Average income by gender (STEM only)
svyby(~ income, ~ as_factor(p02), 
      design = subset(enemdu2023_svydesign, stem == 1), 
      svymean, na.rm = TRUE)

(970.5498 - 1206.6674) / 1206.6674

# Average income by gender (non-STEM only)
svyby(~ income, ~ as_factor(p02), 
      design = subset(enemdu2023_svydesign, stem == 0), 
      svymean, na.rm = TRUE)

(415.4917 - 479.6023) / 479.6023

# Median income of women (STEM only)
women_income <- enemdu2023 %>% filter(p02 == 2, stem == 1) %>% select(income) %>% pull()
women_weights <- enemdu2023 %>% filter(p02 == 2, stem == 1) %>% select(fexp) %>% pull()
median(rep(women_income, times=women_weights), na.rm = TRUE)

# Median income of women (STEM only)
men_income <- enemdu2023 %>% filter(p02 == 1, stem == 1) %>% select(income) %>% pull()
men_weights <- enemdu2023 %>% filter(p02 == 1, stem == 1) %>% select(fexp) %>% pull()
median(rep(men_income, times=men_weights), na.rm = TRUE)
```
